---
title: "LD Estimation Pipeline from VCF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LD Estimation Pipeline from VCF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
runthis <- requireNamespace("VariantAnnotation", quietly = TRUE) &&
  requireNamespace("updog", quietly = TRUE)

if (runthis) {
  runthis <- packageVersion("updog") >= "2.0.2"
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 4,
  eval = runthis
)
set.seed(1)
```

# Summary

This vignette covers how to

1. Load read-count data from a VCF file into R using `VariantAnnotation`,
2. Format these data for genotyping using `updog`,
3. Format the `updog` output for LD estimation using `ldsep`, and
4. Estimate LD using `ldsep` on the formatted output.

We will need the following packages:
```{r setup, message=FALSE}
library(VariantAnnotation)
library(updog)
library(ldsep)
```

If you do not have these packages installed, you can install them in R using:
```{r, eval = FALSE}
install.packages(c("ldsep", "updog", "BiocManager"))
BiocManager::install("VariantAnnotation")
```

The following code chunks are only evaluated if all of these packages are installed. In particular, you need to make sure that your version of `updog` is at least 2.0.2.

```{r}
packageVersion("updog")
```

# Analysis

## Extract data from a VCF file using `VariantAnnotation`

The `ldsep` package comes with a VCF file containing a small (arbitrarily chosen) subset of the data from Uitdewilligen et. al. (2013). These data consist of a diversity panel of autotetraploid *Solanum tuberosum* (2n = 4x = 48). The location of this file on your computer can be found with:
```{r}
uit_file <- system.file("extdata", "subuit.vcf", 
                        package = "ldsep",
                        mustWork = TRUE)
```

We will use the excellent `VariantAnnotation` package (Obenchain et. al., 2014) to load these data into R. You can read more about `VariantAnnotation` in its vignette: <https://bioconductor.org/packages/release/bioc/vignettes/VariantAnnotation/inst/doc/VariantAnnotation.pdf>
```{r}
subuit <- readVcf(file = uit_file)
class(subuit)
```

A brief description of the genotype fields can be seen with
```{r}
geno(header(subuit))
```

See the Samtools website for a more detailed description of the format and the genotype fields in a VCF file: <http://samtools.github.io/hts-specs/>

To format these data for genotyping in `updog`, we need the total read-depth, which in this VCF file is the `DP` field, and we need the read-depth for the reference allele, which in this VCF file is the `RA` field. You should read the [VCF documentation](http://samtools.github.io/hts-specs/) (and the documentation of the software that created the VCF) very carefully, because you might need to work with other genotype fields to obtain read-counts. For example, in many VCF files you need to work with the `AD` field that contains the read-depth for each allele.

```{r}
sizemat <- geno(subuit)$DP
refmat <- geno(subuit)$RA
```

These are matrices where the rows index the loci and the columns index the individuals. That is, `sizemat[i, j]` is the total read-count at SNP `i` for individual `j`, while `refmat[i, j]` is the reference read-count at SNP `i` for individual `j`.

## Genotype individuals using `updog`

To genotype these individuals, we will use the `multidog()` function from the `updog` R package (Gerard et. al., 2018; Gerard and Ferrão, 2020). We could use other genotyping programs, as long as they return genotype log-likelihoods.

We will use `model = "norm"`.
```{r}
ploidy <- 4
mout <- multidog(refmat = refmat, 
                 sizemat = sizemat, 
                 ploidy = ploidy, 
                 model = "norm")
```

We can plot some of the SNPs in a genotype plot (Gerard et. al., 2018) to evaluate the accuracy of the genotyping:
```{r, warning=FALSE, fig.show="hold", out.width="25%", results='hide', fig.alt="Genotype plot of a SNP"}
plot(mout, indices = c(1, 14, 19))
```

We probably want to filter out SNPs that are mostly monoallelic. We can do that with `filter_snp()`.

```{r}
msub <- filter_snp(x = mout, pmax(Pr_0, Pr_1, Pr_2, Pr_3, Pr_4) < 0.95)
nrow(msub$snpdf)
```

We can obtain the genotype log-likelihoods, in the appropriate format used in `ldesp`, with `format_multidog()`.

```{r}
varnames <- paste0("logL_", 0:ploidy)
varnames
larray <- format_multidog(x = msub, varname = varnames)
class(larray)
dim(larray)
```

This is a three-way array of genotype *log*-likelihoods. The first dimension indexes the SNPs, the second dimension indexes the individuals, and the third dimension indexes the genotypes. That is, `larray[i, j, k]` is the genotype log-likelihood at SNP `i` for individual `j` and dosage `k - 1`.

It is possible to also just use the posterior means for LD estimation (doing so will result in speed gains but a loss of accuracy). You can extract the posterior means also using `format_multidog()`.

```{r}
pmmat <- format_multidog(x = msub, varname = "postmean")
class(pmmat)
dim(pmmat)
```

This is a matrix of posterior mean genotypes. That is, `pmmat[i, j]` is the posterior mean genotype at SNP `i` for individual `j`.

## Extracting Log-likelihoods using `fitPoly`

We will also demonstrate how to use the genotyping program, `fitPoly` (Voorrips et. al., 2011; Zych et. al., 2019), to obtain genotype log-likelihoods. None of these code chunks are run by default because I did not want to include more dependencies that are just used in the vignette. You can run these code chunks manually. Makes sure you have the `fitPoly` and `reshape2` packages installed:

```{r, eval=FALSE}
# install.packages(c("fitPoly", "reshape2"))
library(fitPoly)
library(reshape2)
```

We will first need to format the VCF data appropriately for use in `fitPoly`.
```{r, eval = FALSE}
ratiodf <- melt(data = refmat / sizemat, 
                varnames = c("MarkerName", "SampleName"), 
                value.name = "ratio")
saveMarkerModels(ploidy = ploidy, 
                 data = ratiodf, 
                 filePrefix = "uit",
                 rdaFiles = TRUE)
```

This has created two files in your working directory:
    
1. uit_scores.RData: This contains information for each sample. This "RData" file contains a data frame called `scores`.
2. uit_models.RData: This contains information for each SNP. This "RData" file contains a data frame called `modeldata`.
    
We'll load these data frames into R now.
```{r, eval=FALSE}
load("./uit_scores.RData")
load("./uit_models.RData")
```

We will again filter out SNPs that are monoallelic:
```{r,eval=FALSE}
modeldata <- subset(x = modeldata, subset = pmax(P0, P1, P2, P3, P4) < 0.95)
scores <- scores[scores$MarkerName %in% modeldata$MarkerName, ]
```

`fitPoly` only provides posterior genotype probabilities (in the `scores` data frame), not genotype likelihoods. But we can divide these posterior probabilities by the provided prior probabilities (in the `modeldata` data frame) and use these as genotype likelihoods. They aren't quite genotype likelihoods, but they are up to a proportional constant that does not depend on any parameters, and so would result in the same MLEs as using the real genotype likelihoods. This does create a little bit more work for us, formatting-wise, however. 
```{r, eval=FALSE}
mergedf <- merge(x = scores[c("MarkerName", "SampleName", paste0("P", 0:ploidy))], 
                 y = modeldata[c("MarkerName", paste0("P", 0:ploidy))], 
                 by = "MarkerName", 
                 all.x = TRUE)
for (i in 0:ploidy) {
  current_post <- paste0("P", i, ".x")
  current_prior <- paste0("P", i, ".y")  
  current_ll <- paste0("logL_", i)
  mergedf[[current_ll]] <- log(mergedf[[current_post]]) - log(mergedf[[current_prior]])
}

fp_llarray <- acast(
  melt(data = mergedf[c("MarkerName", "SampleName", paste0("logL_", 0:ploidy))],
       id.vars = c("MarkerName", "SampleName"), 
       variable.name = "dosage", 
       value.name = "logl"),
  formula = "MarkerName ~ SampleName ~ dosage", 
  value.var = "logl",
)
```

We can now, if we choose, use `fp_llarray` instead of `larray` in what follows. That is, `fp_llarray[i, j, k]` is the genotype log-likelihood at SNP `i` for individual `j` and dosage `k - 1`. 

Note that for sequencing data, `updog` will generally perform better than `fitPoly`, as `fitPoly` was designed for microarray data. But for microarray data, `fitPoly` works really well.

## Estimate LD using `ldsep`

We will now demonstrate how to estimate LD using the `ldsep` package (Gerard, 2021). To estimate LD using maximum likelihood, directly from genotype log-likelihoods, we will use `mldest()` with the array of genotype log-likelihoods.

```{r}
like_ld <- mldest(geno = larray, K = ploidy, type = "comp")
```

We can visualize the LD estimates using `plot()`
```{r, out.width="50%", fig.alt="Heat map of r-squared"}
plot(like_ld)
```

The naive LD estimates (using just sample moments) with the posterior mean can also be found using `mldest()`:

```{r}
mom_ld <- mldest(geno = pmmat, K = ploidy, type = "comp")
```

The larger naive LD estimates are noticeably attenuated toward zero (due to the influence of genotype uncertainty).
```{r, out.width="50%", fig.alt="Heat map of r-squared"}
plot(mom_ld)
```

It is easier to see this attenuation in a scatterplot. The red, dashed line is the y=x line and any points above this line are pairs of SNPs where the LD was estimated to be larger by the MLE method compared to the naive method. The larger LD values are much more attenuated using the naive approach.
```{r, fig.alt="Scatterplot of naive r-squared versus MLE r-squared"}
par(mar = c(2.4, 2.8, 0, 0) + 0.5, mgp = c(1.8, 0.6, 0))
plot(mom_ld$r2, like_ld$r2, 
     xlab = expression(paste(textstyle(Naive), ~~hat(r)^2)), 
     ylab = expression(paste(textstyle(MLE), ~~hat(r)^2)), 
     pch  = 20)
abline(0, 1, lty = 2, col = 2)
```


Finally, these LD estimates may be obtained in matrix form using `format_lddf()`.

```{r}
ldmat <- format_lddf(obj = like_ld, element = "r2")
ldmat[1:4, 1:4]
```


# References

-  Gerard, David (2021). "Pairwise Linkage Disequilibrium Estimation for Polyploids." _Molecular Ecology Resources_, *21*(4), 1230--1242. [doi:10.1111/1755-0998.13349](https://doi.org/10.1111/1755-0998.13349).

- Gerard D, Ferrão L (2020). "Priors for Genotyping Polyploids." _Bioinformatics_, *36*(6), 1795-1800. ISSN 1367-4803, [doi:10.1093/bioinformatics/btz852](https://doi.org/10.1093/bioinformatics/btz852).

- Gerard D, Ferrão L, Garcia A, Stephens M (2018). "Genotyping Polyploids from Messy Sequencing Data." _Genetics_, *210*(3), 789-807. ISSN 0016-6731, [doi:10.1534/genetics.118.301468](https://doi.org/10.1534/genetics.118.301468).

- Obenchain V, Lawrence M, Carey V, Gogarten S, Shannon P, Morgan M (2014). "VariantAnnotation: a Bioconductor package for exploration and annotation of genetic variants." _Bioinformatics_, *30*(14), 2076-2078. [doi:10.1093/bioinformatics/btu168](https://doi.org/10.1093/bioinformatics/btu168).

- Uitdewilligen JGAML, Wolters AA, D'hoop BB, Borm TJA, Visser RGF, and van Eck HJ (2013). "A next-generation sequencing method for genotyping-by-sequencing of highly heterozygous autotetraploid potato." _PloS one_ *8*(5), e62355. [doi:10.1371/journal.pone.0062355](https://doi.org/10.1371/journal.pone.0062355)

- Voorrips RE, Gort G, and Vosman B (2011). "Genotype calling in tetraploid species from bi-allelic marker data using mixture models." _BMC Bioinformatics_, *12*(1), 172. [doi:10.1186/1471-2105-12-172](https://doi.org/10.1186/1471-2105-12-172).

- Zych K, Gort G, Maliepaard CA, Jansen RC, and Voorrips RE (2019). "FitTetra 2.0–improved
genotype calling for tetraploids with multiple population and parental data support." _BMC bioinformatics_, *20*(1), 148. [doi:10.1186/s12859-019-2703-y](https://doi.org/10.1186/s12859-019-2703-y).
